---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FstArray

<!-- badges: start -->
<!-- badges: end -->

The FstArray R-package provides a DelayedArray array backend for the .fst file format. This includes FstArray for matrix-like data (multiple columns of the same type) and FstColumnVector for data.frame-like data (columns of different types).

FstArray only supports reading from .fst-files. The FstFile-class provides functionality for reading, writing and quering .fst-file metadata.

See the `DelayedArray`, `HDF5Array` and `SQLDataFrame` packages for more information.

## Getting the package

Install from github using:

``` r
devtools::install_github("MalteThodberg/FstArray")
```

Load the package:
```{r load, message=FALSE}
library(FstArray)
```

## Quick usage

IO and introspection with the `FstFile`-class:

```{r FstFile}
# Simulate large data.frame
large_df <- DataFrame(A=rnorm(1e6),
                      B=runif(1e6),
                      C=rpois(1e6, 1),
                      D=sample(letters, 1e6, replace=TRUE),
                      E=factor(sample(LETTERS, 1e6, replace=TRUE)))
object.size(large_df)

# Export
fst_fname <- tempfile(fileext=".fst")
export(large_df, fst_fname)

# Import everything
import(fst_fname)

# Import subset of rows and columns
import(fst_fname, columns=c("A", "E"), from=100, to=150)

# Introspection
ff <- FstFile(fst_fname)
resource(ff)
dim(ff)
colnames(ff)
type(ff)
```

Use .fst files to store array data (all columns of the same type) with the `FstArray`-class:

```{r FstArray}
# FstArray doesn't read the data from disk until needed
fa <- FstArray(fst_fname)

# Resulting object is small, as data is only read from disk when needed.
object.size(fa)

# The type of the resulting array can be directly controlled
fa_int <- FstArray(fst_fname, type="integer")

# The default is to use all numeric and integer columns, but this can be manually specified
fa_char <- FstArray(fst_fname, columns=c("D", "E"))

# All DelayedArray operations are supported for FstArray
fa + 1
rowSums(fa[100:110,])
# See DelayedArray for more
```

Use .fst files to store data.frame data (columns of different types) using the `FstColumnVector`-class:

```{r FstColumnVector}
# FstColumnVectors
fcvs <- FstColumnVectors(fst_fname)

# Delayed operations are supported for each column, causing it to be demoted to a 1D DelayedArray tracking the delayed operations.
fcvs |> 
    subset(D == "a") |> 
    transform(F = A+B) |>
  transform(G = as.vector(F)) # Force delayed operations using as.vector
```
